Joint training of classification model and similarity model for low-resource text classification in chatbot

### 2020-08-12 chatbot_classification/aaai_2021/paper.tex
modification based on ver_1 and update experimental results

\section{Methodology}
In this section, we will describe a system of similarity model fused with classification model, named as SFC, in detail. We come up with 2 different approaches for implementation of SFC system, in which hierarchical SFC is a further improvement of 2-stage SFC for better performance. 
\subsection{2-stage SFC with multi-task learning}
The detailed network structure of 2-stage SFC is shown in Fig. \ref{fig:ver.1.3}. It is composed with 2 separate stages in charge of different jobs.
In stage 1, we can use any auxiliary model or tool to provide top-K most related classes for sampling sentence pairs of both positive samples and negative samples for stage 2. Suppose we have a training data set $\mathcal{D}=\{x_{i},y_{i}\}_{i=1}^{N}$ of $N$ data points, in which $x_{i}$ is the user input query and $y_{i}$ is a single class label from a class label set of $C$ classes in total. If we want to generate sentence pair data set $\mathcal{D'}=\{[(x,x')_{j},l_{j}]\}_{j=1}^{M}$, where $(x, x')_{j}$ is sentence pair grouped by 2 user input queries and $l_{j}\in\{0, 1\}$ is similarity class label, see Equ. \ref{eq:similarity label}, 
\begin{align}
l_{j}=\begin{cases} \label{eq:similarity label}
0 & y_{x}\neq y_{x'}\\
1 & y_{x}=y_{x'}
\end{cases}
\end{align}
directly from original data set $\mathcal{D}$ without any sampling strategy, then $\mathcal{D'}$ will have $M=N^2$ data points of sentence pairs, which might be too time-consuming for the training process of stage 2 since we often have hundreds of class labels under task-specific chatbot scenario. Although we may only have a few data points for a single class label, $N$ is still quite a big number.

Therefore, we adopt the idea of adversarial negative sampling strategy raised by \cite{bamler2020extreme}. The key idea of this sampling strategy is to train with negative samples that are hard to be distinguished from positive samples. Now suppose we have a user input query $x_{i}$ and top-K most related class labels set $\mathcal{C'}=\{c'_{1}, \dots, c'_{K}\}$ found by auxiliary model or tool in stage 1, where $K$ is a hyperparameter set by us that controls candidate class number, we apply the adversarial negative sampling strategy into our sentence pair sampling process by the following 2 steps:

\textbf{1. positive sentence pair sampling}: During the training process, we have to first make sure that the ground truth class label $y_{i}$ for $x_{i}$ is within the candidate class label set $\mathcal{C'}$. If $y_{i}\notin \mathcal{C'}$, we will manually add $y_{i}$ into $\mathcal{C'}$ by replacing $c'_{K}$ with $y_{i}$. Afterwards, we will randomly sample $P$ sentences $\{x'_{0},x'_{1}, \dots, x'_{P}\}$ with the same class label as $y_{i}$ from data set $\mathcal{D}$ to form the set of sentence pairs with positive label $\mathcal{P'}_{i}=\{[(x_{i}, x'_{0}), 1], [(x_{i}, x'_{1}), 1], \dots, [(x_{i}, x'_{P}), 1]\}$, where $P$ is also a hyperparameter set by us that controls the number of sentences we should sample from each class.

\textbf{2. negative sentence pair sampling}: As for negative sentence pairs, we will also randomly sample $P$ sentences for each class in the negative candidate class label set $\mathcal{C'}\backslash \{y_{i}\}$. The class labels in $\mathcal{C'}\backslash \{y_{i}\}$ are the set of most confusing classes comparing to the ground truth ${y_{i}}$, so we assume that the sentence pairs with negative label grouped by user input query $x_{i}$ and sentences with class labels in $\mathcal{C'}\backslash \{y_{i}\}$ are also strong adversarial negative samples for sentence-pair similarity model that can help enhance the training speed and performance. According to the same method as positive sentence pair sampling, we can obtain the set of sentence pairs with negative label as $\mathcal{N'}_{i}=\{[(x_{i}, x'_{0}), 0], [(x_{i}, x'_{1}), 0], ..., [(x_{i}, x'_{P\cdot (K-1)}), 0]\}$

@@

\subsubsection{stage 1: classification model for providing top-K classes}
As for stage 1, our SFC system can adapt to any auxiliary model or even searching tool based on Term Frequency-Inverse
Document Frequency (TF-IDF) retrieval. In this paper, however, we choose to fine-tune RoBERTa on our task-specific data set $\mathcal{D}$ by adding a softmax classifier on top of it, the same as the structure shown in Fig. \ref{Fig.1.sub.1}, for better overall performance. According to our experimental result, RoBERTa can guarantee that approximately 95\% ground truth label be included in top-5 candidate classes with highest probability after being fine-tuned on task-specific data set with experimental result measured by F-score at around 0.75. So we use RoBERTa as encoder of classification model in our setting at stage 1 to ensure low limitation for the performance of stage 2.

@@

Before we start to fine-tune our sentence-pair model on task-specific dataset, we first fine-tune RoBERTa on Quora dataset\cite{iyer2017first}, which contains 404,290 potential duplicate question pairs, for transfer learning. Then we continue to do multi-task training on sentence-pair dataset $\mathcal{D'}$ sampled from the top-K candidate classes provided by stage 1.

@@

As for top-K classification task, we share the same sentence-pair similarity model structure with task 1. The only difference is that we add a task-specific average pooling layer, shown in Fig. \ref{fig:ver.1.3} to implement the classification task based on sentence-pair model.

We already know that the size of ${\Phi}^S$ in task 1 is $M\times 2=(N\cdot K\cdot P)\times 2$, where $N$ is the total number of data points in original training data $\mathcal{D}$, $K$ is a hyperparameter that controls the number of candidate classes provided by stage 1, and $P$ is also a hyperparameter that controls the number of sentences we should randomly sample from each candidate class. Now, for the average pooling layer, we first reshape ${\Phi}^S$ into the size of $N\times K\times P\times 2$, and then split it into ${\Phi}^{K,pos}$ and ${\Phi}^{K,neg}$, and both of them will have the size of $N\times K\times P$. In this way, ${\Phi}^{K,pos}$ can represent the level of similarity for each sentence pair, and in the mean time, ${\Phi}^{K,neg}$ can represent the level of dissimilarity for each sentence pair. Then, we can do average pooling for each candidate class among the top-K classes as shown in Equ. \ref{eq:average pooling:1} and Equ. \ref{eq:average pooling:2}.

@@

Therefore, to further improve the overall performance of SFC, we proposed a hierarchical joint model structure, shown in Fig. \ref{fig:framework}. In hierarchical SFC, classification model is being placed in lower layer level to dynamically provide top-K candidate classes for the sentence-pair similarity model placed in higher layer level through a top-K pooling layer. In this way, classification model and similarity model can be merged into one single joint-model for multi-task training, thus avoiding the limitation brought by 2 separate stage structure.

@@

\section{Experiments}
In this section, we conduct extensive experiments over 5 task-specific dataset to evaluate the performance of our proposed SFC system and the effectiveness of the methodology of multi-task learning and hierachical joint-model structure under low-data settings.

\subsection{Datasets}
The dataset used in our experiments aims to help us study low-resource text classification problem under task-specific conversational chatbot scenario. Therefore, we keep all the dataset under low-resource chatbot corpus setting, which is a large number of class labels (50-250) with a few examples(5-20) in short text for each class, by sampling smaller subsets from the full scale when the raw datasets are oversized, see Table \ref{tbe:dataset statistic}.

@@

@@

ALBERT-xxlarge & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{$0.1422$} & \multirow{2}*{$0.2057$} & \multirow{2}*{$0.5723$} & \multirow{2}*{$0.5912$} & \multirow{2}*{$0.6601$} & \multirow{2}*{$0.6893$}\tabularnewline

@@

RoBERTa-large & \multirow{2}*{$0.8266$} & \multirow{2}*{$0.8861$} & \multirow{2}*{$0.9023$} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{$0.5425$} & \multirow{2}*{$0.6164$} & \multirow{2}*{$0.6503$} & \multirow{2}*{$0.6729$} & \multirow{2}*{$0.7418$} & \multirow{2}*{$0.6362$}\tabularnewline

@@

2-stage SFC & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-}\tabularnewline

@@

\textbf{Group 1: single-task classification model.} This group of baseline models are based on the network structure for classification model shown in \ref{Fig.1.sub.1}. As for the pretrained model, we choose RoBERTa, BERT and ALBERT, all of which are among the most powerful pretrained models that achieve outstanding results in many NLP tasks, in our experimental setting. Since we apply RoBERTa in our SFC system, we naturally adopt single-task classification model based on RoBERTa-large and RoBERTa-base as our baseline models. Besides, we also adopt the baseline pre-trained model BERT-large\cite{devlin2018bert} and recently released ALBERT-xxlarge\cite{lan2019albert} for comparison. Since we use different few-shot settings from \cite{casanueva2020efficient} for those intent detection datasets, we also use BERT-large tuned on our few-shot settings as an essential reference for experimental results.

@@

\textbf{Analysis under few-shot scenarios.} Next, we analyze the overall improvement trend of SFC under various few-shot settings. Our main focus is to study chatbot building in common situation during real-world application where only a few sample sentences are available for each class. The experimental results for the 3 diverse intent detection datasets under few-shot settings indicate that our proposed SFC can achieve +2.47 percents average improvement over one of the most powerful baseline models, which is RoBERTa-large classification model, on all these few-shot settings. Moreover, the improvement in F-score becomes more and more prominent with the decrease of data sample number per class. Our hierarchical SFC can achieve +4.97 percents average improvements over RoBERTa-large classification model in the most extreme few-shot setting with only 5 training samples per class.

\textbf{Fusion of classification and similarity model.} Then, we begin to compare the general performance of SFC with single-task classification and similarity baselines. As for the classification model, hierarchical SFC outperforms BERT-large by +9.81 percents, ALBERT-xxlarge by 4 percents, RoBERTa-large by +2.47 percents, RoBERTa-base by +4.52 percents on average F-score for 5 datasets. In this way, we can see that RoBERTa is the best pre-trained model for the low-resource text classification task we are studying. As for the similarity model, we only choose the strongest pretrained model RoBERTa-large as our baseline, and hierarchical SFC achieves the improvement of 3 percents on average. Moreover, hierarchical SFC's result on 2 datasets without few-shot settings, which are ITG and Amazon-670k, are also 1.24 percents and 2.89 percents higher than Roberta-large classification model. These improvements illustrate that putting classification model and similarity model at different network layers to cooperate with each other is a successful design of hierarchical network structure for joint model, which enables these two models assist each other for better overall results.

>>>>

\section{Methodology}
In this section, we will describe a system of similarity model fused with classification model, named as SFC, in detail. We come up with 2 different approaches for implementation of SFC system, in which hierarchical SFC is a further improvement of 2-stage SFC for better performance. 
\subsection{2-stage SFC with multi-task learning}
The detailed network structure of 2-stage SFC is shown in Fig. \ref{fig:ver.1.3}. It is composed with 2 separate stages in charge of different jobs.
In stage 1, we can use any auxiliary model or tool to provide top-K most related class labels for sampling sentence pairs of both positive samples and negative samples for stage 2. Suppose we have a training data set $\mathcal{D}=\{x_{i},y_{i}\}_{i=1}^{N}$ of $N$ data points, in which $x_{i}$ is the user input query and $y_{i}$ is a single class label from a class label set of $C$ classes in total. If we want to generate sentence pair data set $\mathcal{D'}=\{[(x,x')_{j},l_{j}]\}_{j=1}^{M}$, where $(x, x')_{j}$ is sentence pair grouped by 2 user input queries and $l_{j}\in\{0, 1\}$ is similarity class label, see Equ. \ref{eq:similarity label}, 
\begin{align}
l_{j}=\begin{cases} \label{eq:similarity label}
0 & y_{x}\neq y_{x'}\\
1 & y_{x}=y_{x'}
\end{cases}
\end{align}
directly from original data set $\mathcal{D}$ without any sampling strategy, then $\mathcal{D'}$ will have $M=N^2$ data points of sentence pairs, which might be too time-consuming for the training process of stage 2 since we often have hundreds of class labels under task-specific chatbot scenario. Although we may only have a few data points for a single class label, $M$ is still quite a big number.

Therefore, we adopt the idea of adversarial negative sampling strategy raised by \cite{bamler2020extreme}. The key idea of this sampling strategy is to train with negative samples that are hard to be distinguished from positive samples. Now suppose we have a user input query $x_{i}$ and top-K most related class labels set $\mathcal{C'}=\{c'_{1}, \dots, c'_{K}\}$ provided by auxiliary model or tool in stage 1, where $K$ is a hyperparameter set by us that controls candidate class number, we apply the adversarial negative sampling strategy into our sentence pair sampling process by the following 2 steps:

\textbf{1. positive sentence pair sampling}: During the training process, we have to first make sure that the ground truth class label $y_{i}$ for $x_{i}$ is within the candidate class label set $\mathcal{C'}$. If $y_{i}\notin \mathcal{C'}$, we will manually add $y_{i}$ into $\mathcal{C'}$ by replacing $c'_{K}$ with $y_{i}$, since $c'_{K}$ is the least promising candidate label according to the auxiliary model in stage 1. Afterwards, we will randomly sample $P$ sentences $\{x'_{0},x'_{1}, \dots, x'_{P}\}$ with the same class label as $y_{i}$ from data set $\mathcal{D}$ to form the set of sentence pairs with positive label $\mathcal{P'}_{i}=\{[(x_{i}, x'_{0}), 1], [(x_{i}, x'_{1}), 1], \dots, [(x_{i}, x'_{P}), 1]\}$, where $P$ is also a hyperparameter set by us that controls the number of sentences we should sample from each class.

\textbf{2. negative sentence pair sampling}: As for negative sentence pairs, we will also randomly sample $P$ sentences for each class in the negative candidate class label set $\mathcal{C'}\backslash \{y_{i}\}$. The class labels in $\mathcal{C'}\backslash \{y_{i}\}$ are the set of most confusing class labels comparing to the ground truth ${y_{i}}$, so we assume that the sentence pairs with negative label grouped by user input query $x_{i}$ and sentences with class labels in $\mathcal{C'}\backslash \{y_{i}\}$ are strong adversarial negative samples for sentence-pair similarity model that can help enhance the training speed and performance. According to the same method as positive sentence pair sampling, we can obtain the set of sentence pairs with negative label as $\mathcal{N'}_{i}=\{[(x_{i}, x'_{0}), 0], [(x_{i}, x'_{1}), 0], ..., [(x_{i}, x'_{P\cdot (K-1)}), 0]\}$

@@

\subsubsection{stage 1: classification model for providing top-K candidate class labels}
As for stage 1, our SFC system can adapt to any auxiliary model or even searching tool based on Term Frequency-Inverse
Document Frequency (TF-IDF) retrieval. In this paper, however, we choose to fine-tune RoBERTa on our task-specific data set $\mathcal{D}$ by adding a softmax classifier on top of it, the same as the structure shown in Fig. \ref{Fig.1.sub.1}, for better overall performance. According to our experimental result, RoBERTa can guarantee that approximately 95\% ground truth label be included in top-5 candidate class labels with highest probability after being fine-tuned on task-specific data set with experimental result measured by F-score at around 0.75. So we use RoBERTa as encoder of classification model in our setting at stage 1 to ensure low limitation for the performance of stage 2.

@@

Before we start to fine-tune our sentence-pair model on task-specific dataset, we first fine-tune RoBERTa on Quora dataset\cite{iyer2017first}, which contains 404,290 potential duplicate question pairs, for transfer learning. Then we continue to do multi-task training on sentence-pair dataset $\mathcal{D'}$ sampled from the top-K candidate class labels provided by stage 1.

@@

As for top-K classification task, it shares the same sentence-pair similarity model structure with task 1. The only difference is that we add a task-specific average pooling layer, shown in Fig. \ref{fig:ver.1.3} to implement the classification task based on sentence-pair model.

We already know that the size of ${\Phi}^S$ in task 1 is $M\times 2=(N\cdot K\cdot P)\times 2$, where $N$ is the total number of data points in original training data $\mathcal{D}$, $K$ is a hyperparameter that controls the number of candidate class labels provided by stage 1, and $P$ is also a hyperparameter that controls the number of sentences we should randomly sample from each candidate class labels. Now, for the average pooling layer, we first reshape ${\Phi}^S$ into the size of $N\times K\times P\times 2$, and then split it into ${\Phi}^{K,pos}$ and ${\Phi}^{K,neg}$, and both of them will have the size of $N\times K\times P$. In this way, ${\Phi}^{K,pos}$ can represent the level of similarity for each sentence pair, and in the mean time, ${\Phi}^{K,neg}$ can represent the level of dissimilarity for each sentence pair. Then, we can do average pooling for each candidate class among the top-K classes as shown in Equ. \ref{eq:average pooling:1} and Equ. \ref{eq:average pooling:2}.

@@

Therefore, to further improve the overall performance of SFC, we proposed a hierarchical joint model structure, shown in Fig. \ref{fig:framework}. In hierarchical SFC, classification model is being placed in lower layer level to dynamically provide top-K candidate class labels for the sentence-pair similarity model placed in higher layer level through a top-K pooling layer. In this way, classification model and similarity model can be merged into one single joint-model for multi-task training with sentence pairs sampled from varying candidate class labels, thus avoiding the limitation brought by 2 separate stage structure.

@@

\section{Experiments}
In this section, we conduct extensive experiments over 5 task-specific datasets to evaluate the performance of our proposed SFC system and the effectiveness of the methodology of multi-task learning and hierachical joint-model structure under low-data settings.

\subsection{Datasets}
The datasets used in our experiments aims to help us study low-resource text classification problem under task-specific conversational chatbot scenario. Therefore, we keep all the dataset under low-resource chatbot corpus setting, which is a large number of class labels (50-250) with a few examples(5-20) in short text for each class, by sampling smaller subsets from the full scale when the raw datasets are oversized, see Table \ref{tbe:dataset statistic}.

@@

TextCNN & \multirow{2}*{$0.3539$} & \multirow{2}*{$0.5497$} & \multirow{2}*{$0.6246$} & \multirow{2}*{$0.2882$} & \multirow{2}*{$0.5187$} & \multirow{2}*{$0.6129$} & \multirow{2}*{$0.6554$} & \multirow{2}*{$0.2105$} & \multirow{2}*{$0.3202$} & \multirow{2}*{$0.4154$} & \multirow{2}*{$0.4529$} & \multirow{2}*{$0.5563$} & \multirow{2}*{$0.3333$}\tabularnewline

@@

ALBERT-xxlarge & \multirow{2}*{$0.6500$} & \multirow{2}*{$0.9008$} & \multirow{2}*{$0.9296$} & \multirow{2}*{$0.3608$} & \multirow{2}*{$0.614$} & \multirow{2}*{$0.809$} & \multirow{2}*{$0.853$} & \multirow{2}*{$0.1422$} & \multirow{2}*{$0.2735$} & \multirow{2}*{$0.5723$} & \multirow{2}*{$0.6217$} & \multirow{2}*{$0.7253$} & \multirow{2}*{$0.6893$}\tabularnewline

@@

RoBERTa-large & \multirow{2}*{$0.8266$} & \multirow{2}*{$0.8861$} & \multirow{2}*{$0.9023$} & \multirow{2}*{$0.757$} & \multirow{2}*{$0.8476$} & \multirow{2}*{$0.8614$} & \multirow{2}*{$0.8743$} & \multirow{2}*{$0.5425$} & \multirow{2}*{$0.6164$} & \multirow{2}*{$0.6503$} & \multirow{2}*{$0.6729$} & \multirow{2}*{$0.7418$} & \multirow{2}*{$0.6362$}\tabularnewline

@@

2-stage SFC & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{-} & \multirow{2}*{$0.7997$} & \multirow{2}*{$0.8823$} & \multirow{2}*{$0.8945$} & \multirow{2}*{$0.9123$} & \multirow{2}*{$0.6498$} & \multirow{2}*{$0.6980$} & \multirow{2}*{$0.7202$} & \multirow{2}*{$0.7358$} & \multirow{2}*{$0.8020$} & \multirow{2}*{$0.7311$}\tabularnewline

@@

\textbf{Group 1: single-task classification model.} This group of baseline models are based on the network structure for classification model shown in \ref{Fig.1.sub.1}. As for non-pretrained neural network, we choose TextCNN\cite{kim2014convolutional} as a representative baseline. As for the pretrained model, we choose RoBERTa, BERT and ALBERT, all of which are among the most powerful pretrained models that achieve outstanding results in many NLP tasks, in our experimental setting. Since we apply RoBERTa in our SFC system, we naturally adopt single-task classification model based on RoBERTa-large and RoBERTa-base as our baseline models. Besides, we also adopt the recently released baseline pre-trained model ALBERT-xxlarge\cite{lan2019albert} for comparison. Since we use different few-shot settings from \cite{casanueva2020efficient} for those intent detection datasets, we also use BERT-large tuned on our few-shot settings as an essential reference for experimental results.

@@

\textbf{Analysis under few-shot scenarios.} Next, we analyze the overall improvement trend of SFC under various few-shot settings. Our main focus is to study chatbot building in common situation during real-world application where only a few sample sentences are available for each class. The experimental results for the 3 diverse intent detection datasets under few-shot settings indicate that our proposed SFC can achieve +2.47 percents average improvement over one of the most powerful baseline models, which is RoBERTa-large classification model, on all these few-shot settings. Moreover, the improvement in F-score becomes more and more prominent with the decrease of data sample number per class. Our hierarchical SFC can achieve +4.97 percents average improvements over RoBERTa-large classification model in the most extreme few-shot setting with only 5 training samples per class. This observation indicates that SFC is especially suitable for being applied during the initial stage of building task-specific chatbot system, where the amount of data sample for each class label is extremely scarce.

\textbf{Fusion of classification and similarity model.} Then, we begin to compare the general performance of SFC with single-task classification and similarity baselines. As for the classification model, hierarchical SFC outperforms TextCNN by +37.99 percents, ALBERT-xxlarge by 10 percents, BERT-large by +9.81 percents, RoBERTa-base by +4.52 percents, RoBERTa-large by +2.47 percents on average F-score over 5 datasets. In this way, we can see that RoBERTa is the best baseline pre-trained model for the low-resource text classification task we are studying. As for the similarity model, we only choose the strongest pretrained model RoBERTa-large as our baseline, and hierarchical SFC achieves the improvement of +7.81 percents on average. Moreover, hierarchical SFC's result on 2 datasets without data sampling for few-shot settings, which are ITG and Amazon-670k, are also 1.24 and 2.89 percents higher than Roberta-large classification model, and are 6.96 and 10.77 percents higher than Roberta-large similarity model. These improvements illustrate that putting classification model and similarity model at different network layers to cooperate with each other is a successful design of hierarchical network structure for joint model, which enables these two models assist each other for better overall results.
